Parsing Postfix log files

### 2007-11-18 doc/logparser.tex
Dealt with some more todo items and improved the other parsers section some more.

XXX MOVE plot-* TO A SUBDIRECTORY\@?

@@

    \item Postfix sometimes forcibly kills a \daemon{smtpd} XXX FIGURE OUT
        WHY, so the active connection must be cleaned up, entered in the
        database and deleted from the connections state table.

@@

It is important to compare and contrast newly developed programs,
algorithms and parsers against those already available, to accurately judge
what, if any, improvements are delivered by the newcomers.  There are not
that many previously developed Postfix log parsers, indeed it was quite
difficult to find ten parsers to review for this project, and the
functionality offered ranges from quite basic to much more mature,
depending on the needs of the creator.  The programs are not reviewed from
an independent viewpoint; the objective is to compare and contrast each
program with \parsername{}.

@@

XXX HIGHLIGHT THE DIFFERENCE BETWEEN ADDING NEW RULES AND NEW PARSER
FUNCTIONALITY\@.

XXX NEED A BETTER TERM THAN HARD-CODED, THOUGH THAT MIGHT ARISE FROM THE
PREVIOUS TODO ITEM\@.

@@

        Pflogsumm produces a report designed for troubleshooting, and does
        not support saving any data; parsing is fixed and concentrates on
        extracting only the data required to produce the report.  Both the
        parsing and reporting are difficult to extend, as it is a
        specialised tool.  It does not attempt to correlate log lines by
        queueid or \pid{}, nor does it need to deal with the complications
        encountered during this project.  Pflogsumm produces a useful
        report, and successfully dealt with all the Postfix logs tested
        with.

@@

        Splunk aims to index all an organisation's logs, providing a
        centralised view capable of searching and correlating diverse log
        sources.  The web interface allows complicated searches, providing
        statistics and graphs in real time.  Saved searches can be
        configured to run periodically and the results emailed to a
        recipient or sent to a shell script, which presumably can publish
        the results as required (though possibly without the graphs and
        detailed statistics).  In the specific case of parsing Postfix
        logs, Splunk extracts some standard fields: to and from addresses,
        HELO hostname, date, host the logs were collected from, and
        protocol.  It parses the standard syslog fields at the beginning of
        the line, and extracts any \texttt{key=<value>} pairs; these are
        the fields listed above.  Searches can be made based on the
        extracted fields, and all text in the line is also available for
        searching.  Many types of reports are available, though all are
        variations of a bar or pie chart, with the exception of bubble and
        heatmap graphs.  It is very easy to drill down through the graphs
        to extract a portion of the data (e.g.\ select the hour with the
        largest number of events, then select a particular host, and
        finally a specific address).  It is not possible to search on
        partial words, though the web interface does provide and refine a
        list of possible search terms as you type.  The searches are quite
        fast, though as the free version has a limit on the amount of data
        indexed per day the volume of data to be searched was merely 45mb;
        the cheapest licensed version costs \$6000, and it still limits the
        volume of data indexed per day.

@@

@@

        AWStats will produce simple graphs for many different services, but
        that restricts it to supporting the Lowest Common Denominator.  The
        data it will extract from an \MTA{} log file is limited to:
        \newline \hspace*{2em} time2, email, email\_r, host, host\_r,
        method, url, code and bytesd.  \newline  There does not seem to be
        an explanation of those fields in the documentation.  AWStats has
        no special purpose code to deal with the intricacies of Postfix
        logs, in fact it operates by coercing Postfix logs into
        Apache\footnote{The Apache web server is the most popular HTTP
        server in use for the past 10 years; more information is available
        at \url{http://httpd.apache.org/}} format log files, for analysis
        by AWStats' HTTP log file parser.  The converting parser is hard
        coded, hard to extend, and only deals with a small portion of the
        log lines generated by Postfix, but it does correlate log lines by
        queueid.

@@

        Anteater hasn't been updated since November 2003, and doesn't have
        any English documentation so it's difficult, nigh impossible, to
        accurately comment on what analysis is performs.  It did not run
        successfully when tested, and its parsing would certainly be out of
        date as Postfix has evolved considerably since this tool was last
        updated.  As it neither ran successfully nor has documentation the
        author can read it is difficult to comment on the features it
        provides.

@@

        YAALA uses a plugin based system to analyse log files and produce
        \HTML{} output reports.  The core code is merely 153 lines long, as
        all the parsing and report generation is handled by modules.  YAALA
        supports storing previously gathered data; it dumps its internal
        data structures using Perl's Storable module~\cite{perl-storable},
        so with some knowledge of the data structure created by the module
        responsible for parsing Postfix log files it should be possible to
        extract data with an external program.  Using YAALA as a base would
        be only slightly less work than starting from scratch, as both
        input and output modules would need to be written; it may even be
        more work to implement the parser within the constraints of
        YAALA\@.  YAALA bundles a Postfix parser, which uses hard-coded
        parsing to extract the following fields from specific log lines:

        Aggregations: count(count is not explained), bytes (the sum of
        bytes transferred).\newline Keyfields [sic]: date, hour, sender,
        recipient, defer\_count, delay, incoming\_host, outgoing\_host.

        YAALA provides an \SQL{} like mini-language used when generating
        the reports; sample reports can be seen at~\cite{yaala-samples}.
        Example query: \newline \hspace*{2em} \texttt{requests BY file
        WHERE host =\~{} Google}. \newline The mini-language is quite
        limited, and cannot be used to extract data for external use,
        merely to create reports.

        YAALA as distributed does not attempt the in-depth parsing
        explained in this document, and implementing it within the confines
        of YAALA would be at least as much work as implementing it
        separately.

        In summary YAALA provides a Postfix parser which is not extensible
        unless the user is prepared to change the source
        code,\footnote{\parsername{} requires source code changes for some
        new features, but parsing new log lines is accomplished by adding
        new rules.} and reasonably flexible report generation, but no
        facilities to extract data for use in other tools.

@@

        Lire is a general purpose log parser supporting many different
        types of log file.  Its Postfix parser extracts the following data
        from Postfix logs: \textit{The email servers' reports will show you
        the number of deliveries and the volume of email delivered by day,
        the domains from which you receive or send the most emails, the
        relays most used, etc.\/}; notably rejections are not mentioned.
        It supports multiple output formats (text, \HTML{}, \PDF{} and
        Excel 95).  Logs can be imported and the extracted data saved, and
        reports generated from it later.  Accessing this data from another
        application is undocumented; given the source code it should be
        possible, with time, to access the data.

        Like AWStats and Logrep, Lire attempts to correlate log lines by
        queueid, but not by \pid{}, so the complete list of recipients for
        a mail should be available; however its hard coded parsing extracts
        only part of the available data and makes no attempt to deal with
        the other complications described in
        sections~\refwithpage{complications} and~\refwithpage{additional
        complications}.  It stores data, but not in an easily accessible
        format, and generates reports in multiple formats, but those
        reports do not appear to be customisable.

@@

        Fields are correlated based on the queueid (referred to as
        sessionname [sic] within Logrep), but not by \pid{}.  The parsing
        is hard coded, hard to extend, and error prone: empty fields are
        saved when the log line doesn't match the \regex{}, though it
        appears that they will not overwrite existing data.  It also counts
        the number of log lines parsed and skipped.  Most notably
        rejections are completely ignored, making it unsuitable for the
        purposes of this project.  It doesn't attempt to address any of the
        complications in sections~\refwithpage{complications}
        and~\refwithpage{additional complications} except for correlating
        by queueid.

        Logrep does not come with any documentation, though some scant
        documentation is available on its website.  It requires a web
        browser to interact with it, ensuring that automated log processing
        will be difficult.  Sadly all attempts to use Logrep failed, as it
        was unable to access the log files selected; this appears to be a
        bug rather than operator error.  As parsing failed it wasn't
        possible to review the reports Logrep can generate (available in
        \HTML{} only), nor to examine the (undocumented) format in which it
        can save extracted data for subsequent reuse.

        Logrep extracts far less data from Postfix logs than \parsername{},
        completely ignores rejections, is effectively undocumented, doesn't
        deal with the more complicated aspects of Postfix logs, cannot be
        easily extended by the user, and at the time of writing doesn't
        work properly.

        \url{http://www.itefix.no/phpws/index.php} \newline (Last checked
        2007/10/13.)

@@

While there are other programs available which perform basic Postfix log
parsing (some to a greater level of detail than others), few attempt to
correlate log lines by queueid (none correlate by \pid{}) to produce an
overall record of the journey each mail traverses through Postfix.  None of
the reviewed parsers collect the breadth of information gathered by
\parsername{}, nor make it as easy to extend the parser.  Most other
parsers immediately generate a report and discard the data extracted from
the logs; those which don't discard the data retain it in a format
inaccessible to other tools.  All of the parsers reviewed can produce a
report of greater or lesser detail and complexity, a facility not offered
by \parsername{}; reporting is deferred to a subsequent application.

>>>>

@@

    \item Sometimes an \daemon{smtpd} is killed by a signal, so the active
        connection must be cleaned up, entered in the database and deleted
        from the connections state table.

@@

It is important to compare and contrast newly developed programs,
algorithms and parsers against those already available, to accurately judge
what, if any, improvements are delivered by the newcomers.  There are not
that many previously developed Postfix log parsers, indeed it was quite
difficult to find ten parsers to review for this project, and the
functionality offered ranges from quite basic to much more mature,
depending on the needs of the creator.  The programs are not reviewed from
an independent viewpoint; the objective is to compare and contrast each
program with \parsername{}.  One important difference between \parsername{}
and the parsers reviewed is that only \parsername{} makes it possible to
parse new log lines without modifying the core parsing algorithm.  Although
this may seem like a trivial point, is it substantially more difficult to
understand a program's entire parsing algorithm, identify the correct
location to change, and make the appropriate changes without adversely
affecting parsing, particularly as there may be edge cases which are not
immediately obvious.\footnote{See
section~\refwithpage{yet-more-aborted-delivery-attempts} for a complication
which occurs only four times in 93 log files tested.}  Requiring changes to
the parsing algorithm also complicates upgrades, as the changes must be
preserved, and may clash with changes made by the developer.  \parsername{}
allows the user to add new rules to the database without changing the
parsing algorithm, unless the new log lines to be parsed require
functionality not already provided by the algorithm.  If the new log lines
do require new functionality, new actions can be added to the parser
without modifying existing actions or other parts of the algorithm; only in
the rare case that the new actions require support from other sections of
the code will more extensive changes be required.

XXX ADD THE PREVIOUS TWO SENTENCES TO THE PARSING RULES/ALGORITHM
SECTION\@.

@@

@@

        Pflogsumm produces a report designed for troubleshooting, and does
        not support saving any data; parsing concentrates on extracting
        only the data required to produce the report.  Both the parsing and
        reporting are difficult to extend, as it is a specialised tool.  It
        does not attempt to correlate log lines by queueid or \pid{}, nor
        does it need to deal with the complications encountered during this
        project.  Pflogsumm produces a useful report, and successfully
        dealt with all the Postfix logs tested with.

@@

        Splunk aims to index all an organisation's logs, providing a
        centralised view capable of searching and correlating diverse log
        sources.  The web interface allows complicated searches, providing
        statistics and graphs in real time.  Saved searches can be
        configured to run periodically and the results emailed to a
        recipient or sent to a shell script, which presumably can publish
        the results as required (though possibly without the graphs and
        detailed statistics).  In the specific case of parsing Postfix
        logs, Splunk extracts some standard fields: to and from addresses,
        HELO hostname, date, host the logs were collected from, and
        protocol.  It parses the standard syslog fields at the beginning of
        the line, and extracts any \texttt{key=<value>} pairs; these are
        the fields listed above.  Searches can be based on the extracted
        fields, and all text in the line is also available for searching.
        Many types of reports are available, though all are variations of a
        bar or pie chart, with the exception of bubble and heatmap graphs.
        It is very easy to drill down through the graphs to extract a
        portion of the data (e.g.\ select the hour with the largest number
        of events, then select a particular host, and finally a specific
        address).  It is not possible to search on partial words, though
        the web interface does provide and refine a list of possible search
        terms as you type.  The searches are quite fast, though as the free
        version has a limit on the amount of data indexed per day the
        volume of data to be searched was merely 45mb; the cheapest
        licensed version costs \$6000, and it still limits the volume of
        data indexed per day.

@@

        XXX TEST THIS\@: THE NEXT PARAGRAPH IS CRAP\@.

@@

        AWStats will produce simple graphs for many different services, but
        supporting many different services without special purpose code
        restricts it to supporting the \LCD{}.  The data it will extract
        from an \MTA{} log file is limited to: \newline \hspace*{2em}
        time2, email, email\_r, host, host\_r, method, url, code and
        bytesd.  \newline  There does not seem to be an explanation of
        those fields in the documentation.  AWStats has no special purpose
        code to deal with the intricacies of Postfix logs, in fact it
        operates by coercing Postfix logs into Apache\footnote{The Apache
        web server is the most popular HTTP server in use for the past 10
        years; more information is available at
        \url{http://httpd.apache.org/}} format log files, for analysis by
        AWStats' HTTP log file parser.  The converting parser only deals
        with a small portion of the log lines generated by Postfix, and
        though it does correlate log lines by queueid, it does not deal
        with any of the other complications described in this document.

@@

        Anteater doesn't have any English documentation so it's difficult,
        nigh impossible, to accurately comment on what analysis is
        performs.  It did not run successfully when tested, and its parsing
        would certainly be out of date as Postfix has evolved considerably
        since this tool was last updated (November 2003).  As it neither
        ran successfully nor has documentation the author can read a
        detailed review cannot be provided.

@@

        YAALA uses a plugin based system to analyse log files and produce
        \HTML{} output reports.  The core code is merely 153 lines long, as
        all the parsing and report generation is handled by modules.  YAALA
        supports storing previously gathered data; it dumps its internal
        data structures using Perl's Storable module~\cite{perl-storable},
        so with some knowledge of the data structure created by the module
        responsible for parsing Postfix log files it should be possible to
        extract data with an external program.  Using YAALA as a base would
        be only slightly less work than starting from scratch, as both
        input and output modules would need to be written specially; it may
        even be more work to implement the parser within the constraints of
        YAALA\@.  YAALA bundles a Postfix parser, which extracts the
        following fields from specific log lines:

        \hspace*{2em}Aggregations: count (not explained), bytes (sum of
        bytes transferred).\newline \hspace*{2em}Keyfields [sic]: date,
        hour, sender, recipient, defer\_count, delay, incoming\_host,
        outgoing\_host.

        YAALA provides an \SQL{} like mini-language used when generating
        the reports; sample reports can be seen at~\cite{yaala-samples}.
        Example query: \newline \hspace*{2em} \texttt{requests BY file
        WHERE host =\~{} Google}. \newline The mini-language is quite
        limited and cannot be used to extract data for external use, merely
        to create reports.

        YAALA as distributed does not attempt the in-depth parsing
        explained in this document, and implementing such parsing within
        the confines of YAALA would be at least as much work as
        implementing it separately.

        In summary YAALA provides a Postfix parser which only handles the
        most common Postfix log lines, provides reasonably flexible report
        generation from the limited data extracted, but no facilities to
        extract data for use in other tools.

@@

        Lire is a general purpose log parser supporting many different
        types of log file.  Its Postfix parser extracts the following data
        from Postfix logs: \textit{The email servers' reports will show you
        the number of deliveries and the volume of email delivered by day,
        the domains from which you receive or send the most emails, the
        relays most used, etc.\/}; notably rejections are not mentioned or
        handled.  It supports multiple output formats (text, \HTML{},
        \PDF{} and Excel 95).  Logs can be parsed, the extracted data
        saved, and reports generated from it later.  Accessing this data
        from another application is undocumented; given the source code it
        should be possible, with sufficient time and effort, to access the
        data.

        Like AWStats and Logrep, Lire attempts to correlate log lines by
        queueid, but not by \pid{}, so the complete list of recipients for
        a mail should be available; however its parser extracts only part
        of the available data and makes no attempt to deal with the other
        complications described in sections~\refwithpage{complications}
        and~\refwithpage{additional complications}.  It stores data, but
        not in an easily accessible format, and generates reports in
        multiple formats, but those reports do not appear to be
        customisable.

@@

        Fields are correlated based on the queueid (referred to as
        sessionname [sic] within Logrep), but not by \pid{}.  The parsing
        is error prone: empty fields are saved when the log line doesn't
        match the \regex{}, though it appears that they will not overwrite
        existing data.  It also counts the number of log lines parsed and
        skipped.  Most notably rejections are completely ignored, making it
        unsuitable for the purposes of this project.  It doesn't attempt to
        address any of the complications in
        sections~\refwithpage{complications} and~\refwithpage{additional
        complications} except for correlating by queueid.

        Logrep does not come with any documentation, though some scant
        documentation is available on its website.  It requires a web
        browser to interact with it, ensuring that automated log processing
        will be difficult.  Sadly all attempts to use Logrep failed, as it
        was unable to access the log files selected; this appears to be a
        bug rather than operator error.\footnote{If it is caused by
        operator error, the interface needs improvement as the (minimal)
        instructions were followed as closely as possible.}  As parsing
        failed it wasn't possible to review the reports Logrep can generate
        (available in \HTML{} only), nor to examine the (undocumented)
        format in which it can save extracted data for subsequent reuse.

        Logrep extracts far less data from Postfix logs than \parsername{},
        completely ignores rejections, is effectively undocumented, doesn't
        deal with the more complicated aspects of Postfix logs, and at the
        time of writing doesn't work properly.

        \url{http://www.itefix.no/phpws/index.php} \newline (Last checked
        2007/11/18.)

@@

While there are other programs available which perform basic Postfix log
parsing (some to a greater level of detail than others), few attempt to
correlate log lines by queueid (none correlate by \pid{}) to produce an
overall record of the journey each mail traverses through Postfix.  None of
the reviewed parsers collect the breadth of information gathered by
\parsername{}, nor make it as easy to extend the parser to handle new log
lines.  Most other parsers immediately generate a report and discard the
data extracted from the logs; those which don't discard the data retain it
in a format inaccessible to other tools.  All of the parsers reviewed can
produce a report of greater or lesser detail and complexity, a facility not
offered by \parsername{}; reporting is deferred to a subsequent
application.
